Equation-free analysis of a dynamically evolving multigraph

### 2016-02-08 report/multigraph_manuscript.tex
More paper work

We study the edge-conservative preferential attachment model, a detailed mathematical analysis of which can be found in \cite{rath_time_2009}. The system evolves in discrete steps $t = 0,1,\ldots t_f$, and we denote the $n$-vertex graph at each point by $G_n(t)$. The initial state, $G_n(0)$, is an arbitrary distribution of $m$ edges among the $n$ vertices. No restrictions are placed on this initial distribution: multiple edges and loops are permitted.\ignore{an Erd\H{o}s-R\'{e}nyi random graph on $n$ vertices and $m$ edges. Here, however, $m$ may be larger than the total number of edges in a completely-connected $n$-vertex graph so we cannot assign a probability to each edges existence, instead placing $m$ edges at random in the graph.} The system is then advanced step-by-step based on the following procedure:

@@

Results of the application of this general framework with the specific lifting and restriction operators previously outlined are shown in Fig. (\ref{fig:cpi-results}). We see good agreement between the CPI-accelerated and normal systems, while reducing the number of detailed steps by one third. It is important to mention that this method was applied not to a single system instantiation, but to a large ensemble. This ensured that when properties such as the fitted polynomial coefficients were averaged over the ensemble they evolved smoothly despite the stochastic nature of the system, as shown previously in Fig. (\ref{fig:coeff-evo}).

@@

When applying machine learning techniques to a dataset, it is
necessary to define a distance (or similarity) between each pair of
data point. If these points are a set of vectors in $\mathbb{R}^n$ one
has a number of metrics to choose from, the Euclidean distance being a
common first option. Unfortunately, when individual points are not
vectors but networks, the definition of a computationally feasible
metric becomes far more challenging. Examples such as the maximal
common subgraph and edit distances, defined in \cite{bunke_graph_1998}
and \cite{gao_survey_2010} do define metrics on the space of graphs,
but their computation is $NP\textnormal{-}hard$. Other computationally
feasible approaches include comparing distributions of random
walks on each graph \cite{kondor}, calculating the so-called
$n$-tangle density \cite{fefferman}, or calculting the edit distance
with an approximation algorithm \cite{riesen,zeng}.  \\

The strategy used in the following methods, detailed in
\cite{rajendran_analysis_2013} and \cite{gu} enumerates the number of
times a certain set of motifs (or subgraphs) appears in each network
in the dataset. This maps each network to a vector in $\mathbb{R}^n$,
and the Euclidean distance is subsequently used to calculate the
similarity of two graphs. Due to computational restrictions, we chose
to count the number of three- and four-vertex subgraphs that were
contained in each network. As there are eight such motifs, shown in
Fig. (\ref{fig:motifs}), this process $\gamma$ maps each graph to an
eight-dimensional vector: $\gamma : G \rightarrow
\mathbb{R}^8$. Additionally, we applied this operateration to a
simplified version of each graph, wherein any multiple edges were
reduced to single edges.

@@

\subsection{PCA}

% TODO (referenced fig below)

PCA is used to linearly project data into subspaces that capture the
directions along with the data varies most
\cite{jolliffe_principal_2014}. It has found wide application, but
suffers from its restriction to linear embeddings. Indeed, many
datasets will not lie along some hyperplane embedded in
$\mathbb{R}^n$, but rather will form a nonlinear manifold throughout
the space. Thus it is not unexpected that such a method might work
poorly when applied to our nonlinear system. Results shown in
Fig. (\ref{fig:pca}) show that, while worse than the embedding
uncovered by DMAPS in the next section, PCA still offers a reasonable
dimensionality reduction.

@@

Unlike PCA, DMAPS uncovers underlying nonlinear manifolds hidden in data. This is achieved by solving for the discrete equivalent of the eigenfunctions and eigenvalues of the Laplacian operator over the manifold, which amounts to calculating leading eigenvector/eigenvalue pairs of a symmetric Markov matrix $W$. See \cite{coifman_diffusion_2006,nadler_diffusion_2006} for details. Results from embeddings on different timescales are shown in Fig. (\ref{fig:dmaps-results}). We see that on the short, $n^2$ timescale, the differences in initial conditions lead to well-separated embeddings in three-dimensions represented by the small spheres in Fig. (\ref{fig:dmaps-results}), while at long times, we see that both systems evolve to the same steady state over a shared one-dimensional curve, as would be expected given the one-dimensional parameterization possible in time. \\

@@

The equation free framework was successfully applied to the edge-conservative preferential attachment model, accelerating simulations through CPI and locating stationary states through coarse Newton-GMRES. Additionally the underlying low-dimensional embedding was detected via PCA and DMAPS. \\

An open area of investigation is the interpretation of the output from the PCA and DMAPS techniques. As a linear map, PCA retains a clear relationship between its input and output, but, as we saw, it is less effective when data is highly nonlinear. DMAPS may perform better the task of dimensionality reduction, but it is unclear what the embedding coordinates relate to in the data.

@@

\bibliographystyle{abbrv}
\bibliography{manuscript-refs.bib}

>>>>

We study the edge-conservative preferential attachment model, a detailed mathematical analysis of which can be found in \cite{rath_time_2009} and \cite{rath_multigraph_2012}. The system evolves in discrete steps $t = 0,1,\ldots t_f$, and we denote the $n$-vertex graph at each point by $G_n(t)$. The initial state, $G_n(0)$, is an arbitrary distribution of $m$ edges among the $n$ vertices. No restrictions are placed on this initial distribution: multiple edges and loops are permitted.\ignore{an Erd\H{o}s-R\'{e}nyi random graph on $n$ vertices and $m$ edges. Here, however, $m$ may be larger than the total number of edges in a completely-connected $n$-vertex graph so we cannot assign a probability to each edges existence, instead placing $m$ edges at random in the graph.} The system is then advanced step-by-step based on the following procedure:

@@

Results of the application of this general framework with the specific lifting and restriction operators previously outlined are shown in Fig. (\ref{fig:cpi-results}). We see good agreement between the CPI-accelerated and normal systems, while reducing the number of detailed steps by one third. It is important to mention that this method was applied not to a single system instantiation, but to a large ensemble. This ensured that when properties such as the fitted polynomial coefficients were averaged over the ensemble they evolved smoothly despite the stochastic nature of the system.

@@

When applying machine learning techniques to a dataset, it is
necessary to define a distance (or similarity) between each pair of
data point. If these points are a set of vectors in $\mathbb{R}^n$ one
has a number of metrics to choose from, the Euclidean distance being a
common first option. Unfortunately, when individual points are not
vectors but networks, the definition of a computationally feasible
metric becomes far more challenging. Examples such as the maximal
common subgraph and edit distances, defined in \cite{bunke_graph_1998}
and \cite{gao_survey_2010} do define metrics on the space of graphs,
but their computation is $NP\textnormal{-}hard$. Other computationally
feasible approaches include comparing distributions of random
walks on each graph \cite{vishwanathan_graph_2010}, calculating the so-called
$n$-tangle density \cite{gallos_revealing_2014}, or calculting the edit distance
with an approximation algorithm \cite{riesen_approximate_2009,zeng_comparing_2009}.  \\

The strategy used in the following methods, detailed in
\cite{rajendran_analysis_2013} and \cite{xiao_structure-based_2008}
enumerates the number of times a certain set of motifs (or subgraphs)
appears in each network in the dataset. This maps each network to a
vector in $\mathbb{R}^n$, and the Euclidean distance is subsequently
used to calculate the similarity of two graphs. Due to computational
restrictions, we chose to count the number of three- and four-vertex
subgraphs that were contained in each network. As there are eight such
motifs, shown in Fig. (\ref{fig:motifs}), this process $\gamma$ maps
each graph to an eight-dimensional vector:
$\gamma : G \rightarrow \mathbb{R}^8$. Additionally, we applied this
operateration to a simplified version of each graph, wherein any
multiple edges were reduced to single edges.

@@

\subsection{PCA}
\label{sec:pca}
% TODO (referenced fig below)

PCA is used to linearly project data into subspaces that capture the
directions along which the data varies most
\cite{jolliffe_principal_2014}. It has found wide application, but
can suffer from its restriction to linear embeddings. Indeed, many
datasets will not lie along some hyperplane embedded in
$\mathbb{R}^n$, but rather will form a nonlinear manifold throughout
the space. \\

Theoretical results in \cite{rath_time_2009} state that, for a given network
size $n$, the final stationary state depends only on the number of
edges present $m$ and the model parameter $\kappa$. To assess both the
validity of our graph embedding technique and the usefullness of the
PCA in the context of complex networks, we generated a dataset of
stationary states over a range of $m$ and $\kappa$ values by directly
running the model over $O(n^3)$ steps ($N$ values of each parameter
were taken, for a total number of $N^2$ networks). Each resulting
graph $G(m_i, \kappa_j) = G_{ij}$ was then embedded into
$\mathbb{R}^8$ by counting the number of times each of the subgraphs
shown in Fig. (\ref{fig:motifs}) appeared in the network. Thus
$\gamma(G_{ij}) = v_{ij} \in \mathbb{R}^8$. We then proceeded to
perform a principal component analysis on this collection of vectors
${v_{ij}}_{i,j=1}^N$. Surprisingly, this linear technique succeeded in
uncovering a two-dimensional embedding of the dataset corresponding to
the two underlying parameters $m$ and $\kappa$, as shown in
Fig. (\ref{fig:pca}). This suggests that, given some final network
state $G_f(m, \kappa)$, by projecting its embedding onto these first two
principal components one could obtain a reasonable approximation of
the hidden parameter $\kappa$.

@@

Unlike PCA, DMAPS uncovers underlying nonlinear manifolds hidden in data. This is achieved by solving for the discrete equivalent of the eigenfunctions and eigenvalues of the Laplacian operator over the manifold, which amounts to calculating leading eigenvector/eigenvalue pairs of a symmetric Markov matrix $W$. See \cite{coifman_diffusion_2006,nadler_diffusion_2006} for details. \\

First we applied DMAPS to the dataset described in
Sec. (\ref{sec:pca}). Given the relative success of PCA in this
setting, one would expect DMAPS to also uncover the two-dimensional
embedding corresponding to different values of $m$ and $\kappa$. As
Fig. (\ref{fig:dmaps-rk}) shows, this is indeed the
case. Additionally, we were interested in embeddings of different
model trajectories. This dataset was generated by sampling two
different model simulations at even intervals as they evolved ($N$
points were sampled from each trajectory). The parameter values $m$
and $\kappa$ were held constant, but one graph was initialiazed as an
Erd\H{o}s-R\'{e}nyi random graph (Fig. (\ref{fig:erdos-init})), while
the other was initialized as a lopsided graph
(Fig. (\ref{fig:lopsided-init})). Letting $G_e(t)$ refer to the
Erd\H{o}s-R\'{e}nyi-initialized system at step $t$, and $G_l(t)$ the
lopsided-initialized system, the embedding $\gamma$ was used to create
points $\gamma(G_e(t) = v_e(t) \in \mathbb{R}^8$ and similarly
$\gamma(G_l(t) = v_l(t) \in \mathbb{R}^8$. \\

DMAPS was then applied to this set of $2N$ points, the results of
which are shown in Fig. (\ref{fig:dmaps-results}). At first, the two
points are mapped to distant regions in $\mathbb{R}^3$ due to their
different initial conditions. They also evolve along different
coordinates of the embedding, showing they traverse two distinct
trajectories on their approach to the stationary state. The particular
eigenvector each trajectory travels along is also significant: as
$G_e(t)$ was mapped to $\Phi_1$ and $G_l(t)$ to $\Phi_2$, we can infer
that the dynamics of the Erd\H{o}s-R\'{e}nyi-initialized system occur
more slowly than those of the lopsided-initialized system. Indeed,
referring back to Fig. (\ref{fig:dse}), we see that the lopsided
network descends more rapidly to the stationary state. Eventually,
their embeddings meet as they arrive at this final, shared state. Thus
DMAPS has proven useful in elucidating both geometric and dynamic
features of the system as shown in Figs. (\ref{fig:dmaps-rk}) and (\ref{fig:dmaps-results}). \\

@@

The equation free framework was successfully applied to the edge-conservative preferential attachment model, accelerating simulations through CPI and locating stationary states through coarse Newton-GMRES. Additionally the underlying two-dimensional description was uncovered via PCA and DMAPS. \\

An open area of investigation is the interpretation of the output from the PCA and DMAPS techniques. As a linear map, PCA retains a clear relationship between its input and output, but is less effective when data is highly nonlinear. DMAPS may perform better the task of dimensionality reduction, but it is unclear what the embedding coordinates relate to in the data.

@@

\bibliographystyle{abbrv}
\bibliography{multigraph-refs.bib}
